---
title: "FCB enrollment"
author: "EMB"
format:
  dashboard:
    self-contained: true
editor: visual
---

# FCB enrollment by cohort

```{r}
load(here::here("_outputs/Rdata/data_image1.RData"))
# load packages needed by ALL chunks
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(scales)
library(reactable)
library(stringr)
library(gt)
library(forcats)   # << this was missing

make_heat_gt <- function(df, title, key_col = NULL,
                         palette = c("white", "steelblue")) {

  # Normalize to a data frame with an explicit Cohort column
  if (!is.null(key_col) && key_col %in% names(df)) {
    df2 <- df
    names(df2)[names(df2) == key_col] <- "Cohort"
  } else {
    if (is.null(rownames(df))) stop("Provide key_col when df has no rownames.")
    df2 <- tibble::rownames_to_column(as.data.frame(df), var = "Cohort")
  }

  # Split body rows vs. Total row for computing the color scale
  body_df   <- df2 %>% filter(Cohort != "Total")
  total_row <- df2 %>% filter(Cohort == "Total")

  # Numeric columns to color
  num_cols <- names(body_df)[sapply(body_df, is.numeric)]
  # Remove the first four columns/terms from the numeric columns
  if (length(num_cols) > 4) {
  drop_cols <- num_cols[1:4]
  df2 <- df2 %>% select(-all_of(drop_cols))
  num_cols <- setdiff(num_cols, drop_cols)
}
  
  # Safe domain from body rows only
  rng <- range(as.matrix(body_df[, num_cols, drop = FALSE]), na.rm = TRUE)
  if (!is.finite(rng[1]) || !is.finite(rng[2])) rng <- c(0, 1)

  # Build gt from the full table (keeps Total visible)
  gt(df2, rowname_col = "Cohort") %>%
    fmt_number(columns = everything(), decimals = 0, sep_mark = ",") %>%
    sub_missing(columns = everything(), missing_text = "") %>%
    # Color only non-Total rows, with domain from non-Total rows
    data_color(
      columns = all_of(num_cols),
      rows    = Cohort != "Total",
      colors  = col_numeric(palette = palette, domain = rng, na.color = "transparent")
    ) %>%
    # Emphasize forecast columns
    tab_style(
      style = list(cell_text(style = "italic", weight = "bold")),
      locations = cells_body(columns = any_of(c("F25", "S26")))
    ) %>%
    # Optional: visually emphasize the Total row (no coloring already)
    tab_style(
      style = list(cell_text(weight = "bold")),
      locations = cells_body(rows = Cohort == "Total")
    ) %>%
    tab_header(title = title) %>%
    opt_row_striping()
}

```

::: {layout-ncol="2"}
```{r}
gt_pre <- make_heat_gt(fcb_pre_df, title = "FCB pre-major enrollment (by cohort)")
gt_pre
```

```{r}
gt_ud  <- make_heat_gt(fcb_ud_df,  title = "FCB UD enrollment (by cohort)")
gt_ud
```
:::

Tables show FCB pre-major (left) and upper-division (right) headcounts by term and by admissions cohort. Rows represent First-time (FT) and Transfer (TR) cohorts by year of entry. The columns for F25 and S26 are ***forecasts*** from a non-homogeneous Markov chain model using historical transition probabilities for major migration (or "internal transfer"), attrition, graduation, and promotion to upper-division status. (Data source: ASIR Cohort Analysis Report Builder, Enrollment Services reports on F25 incoming cohorts.) Use the "Expand" button at lower-right to view full tables.

# Totals

```{r}
fcb_tot_df <- fcb_pre_df + fcb_ud_df 
gt_tot <- make_heat_gt(fcb_tot_df, title = "FCB total enrollment (by cohort)")
gt_tot
```

# Waterfall table

```{r}
# 1) Prepare (order columns, tidy percent)
tbl <- fcbud_waterfall %>%
  mutate(
    Semester = as.integer(Semester),
    Group = ifelse(str_starts(Cohort, "FT"), "FT cohorts", 
                   ifelse(Cohort == "Total", "Totals", "TR cohorts")),
    Pct_of_Total = Pct_of_Total/100  # convert to proportion if it's in %
  ) %>%
  select(Group, Cohort, Semester, F23, F24, Diff, Pct_of_Total, A, B, C, D, E)

# 2) Build a handsome gt table
gt(tbl, groupname_col = "Group") %>%
  tab_header(
    title = md("**FCB upper-division headcount change: F23 → F24**"),
    subtitle = "Waterfall rows with factor decomposition"
  ) %>%
  tab_spanner(label = "Factors (Δ decomposition)", columns = c(A, B, C, D, E)) %>%
  cols_label(
    Cohort = "Cohort (pair)",
    Semester = "Semester",
    F23 = "F23",
    F24 = "F24",
    Diff = "Δ",
    Pct_of_Total = "% of total",
    A = "Admissions N", B = "% FCB", C = "Retention",
    D = "Internal Transfers", E = "Promotion"
  ) %>%
  # Add this line to center columns (including headers and values)
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  fmt_number(columns = c(F23, F24, Diff, A:E), decimals = 0, use_seps = TRUE) %>%
  fmt_percent(columns = Pct_of_Total, decimals = 1) %>%
  data_color(
    columns = Diff,
    colors = scales::col_bin(
      palette = c("#d95f02","#f0f0f0","#1b9e77"),
      domain = c(min(tbl$Diff, na.rm=TRUE), max(tbl$Diff, na.rm=TRUE)),
      bins = c(-Inf, 0, Inf)
    )
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_row_groups(groups = "Totals")
  ) %>%
  tab_options(table.font.size = px(14)) %>%
  opt_row_striping()
```

This table breaks down the growth in FCB upper-division enrollment between Fall 2023 and Fall 2024 into contributing factors, including admissions size and FCB composition, retention and graduation, internal transfers, and promotion to upper-division status. Each row shows how much each factor contributed to the headcount change of +501: a row sum of the Factors yields the color-coded Δ. (Data source: ASIR Cohort Analysis Report Builder.) *Want more details? \[See the full method.\]*

# Waterfall visual

```{r}
## Classic Waterfall F23 --> F24

library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(scales)
library(reactable)
library(stringr)
library(gt)
library(forcats)

make_waterfall_topk_df <- function(fcbud_waterfall,
                                   term1 = "F23",
                                   term2 = "F24",
                                   top_k = 10,
                                   order_by = c("abs_diff","pct","diff"),
                                   verbose = TRUE) {
  order_by <- match.arg(order_by)
  
  # Split rows
  has_total_row <- any(fcbud_waterfall$Cohort == "Total")
  cohorts_only  <- fcbud_waterfall %>% filter(Cohort != "Total")
  
  # Totals (use Total row if present)
  if (has_total_row) {
    total_t1 <- fcbud_waterfall[[term1]][fcbud_waterfall$Cohort == "Total"]
    total_t2 <- fcbud_waterfall[[term2]][fcbud_waterfall$Cohort == "Total"]
  } else {
    total_t1 <- sum(cohorts_only[[term1]], na.rm = TRUE)
    total_t2 <- sum(cohorts_only[[term2]], na.rm = TRUE)
  }
  total_delta <- total_t2 - total_t1
  
  # Build a lookup of Cohort → Semester for labeling
  sem_lookup <- fcbud_waterfall %>%
    filter(Cohort != "Total") %>%
    distinct(Cohort, Semester)
  
  # Order (with % fallback)
  core <- cohorts_only %>%
    mutate(
      pct_used  = ifelse(is.na(Pct_of_Total), 100 * Diff / total_delta, Pct_of_Total),
      order_key = case_when(
        order_by == "abs_diff" ~ abs(Diff),
        order_by == "pct"      ~ abs(pct_used),
        TRUE                   ~ Diff
      )
    ) %>%
    arrange(desc(order_key)) %>%
    # ensure Semester is present (if your table already has it, this is harmless)
    left_join(sem_lookup, by = "Cohort")
  
  n_avail <- nrow(core)
  top_k   <- min(top_k, n_avail)
  
  # Build a lookup: Cohort -> Semester (from the already-built table)
  sem_map <- with(
    fcbud_waterfall %>% dplyr::filter(Cohort != "Total"),
    stats::setNames(Semester, Cohort)
  )
  
  # Top-K with (S#) labels via the lookup (no joins, no ambiguity)
  top <- core %>%
    dplyr::slice_head(n = top_k) %>%
    dplyr::mutate(
      Sem_for_label = sem_map[Cohort],                 # vectorized lookup
      label = ifelse(
        is.na(Sem_for_label),
        Cohort,                                       # fallback if missing
        paste0(Cohort, " (S", as.integer(Sem_for_label), ")")
      )
    ) %>%
    dplyr::transmute(
      step = Cohort,
      label = label,
      value = Diff,
      type  = ifelse(Diff >= 0, "Increase", "Decrease")
    )
  
  
  # “Other” to force Start + steps = End (full)
  other_sum <- total_delta - sum(top$value, na.rm = TRUE)
  contrib <- if (abs(other_sum) > 0) {
    bind_rows(
      top,
      tibble(step = "Other", label = "Other", value = other_sum,
             type = ifelse(other_sum >= 0, "Increase", "Decrease"))
    )
  } else top
  
  if (verbose) {
    message(sprintf("Top-K selected: %d (of %d). 'Other' = %+d. Sum(top + other) = %+d. Full Δ = %+d.",
                    top_k, n_avail, round(other_sum),
                    round(sum(contrib$value, na.rm = TRUE)), round(total_delta)))
  }
  
  # Build sequence: Start → contributions → End
  seq_df <- bind_rows(
    tibble(step = "Start", label = term1, value = total_t1, type = "Start"),
    contrib,
    tibble(step = "End",   label = term2, value = total_t2, type = "End")
  )
  
  # Floating bases via running total
  running <- total_t1
  ymin <- numeric(nrow(seq_df)); ymax <- numeric(nrow(seq_df))
  for (i in seq_len(nrow(seq_df))) {
    tp  <- seq_df$type[i]; val <- seq_df$value[i]
    if (tp == "Start") { ymin[i] <- 0; ymax[i] <- total_t1; running <- total_t1 }
    else if (tp %in% c("Increase","Decrease")) { ymin[i] <- running; running <- running + val; ymax[i] <- running }
    else if (tp == "End") { ymin[i] <- 0; ymax[i] <- total_t2 }
  }
  
  seq_df$ymin <- ymin; seq_df$ymax <- ymax
  seq_df$x    <- seq_along(seq_df$step)
  seq_df$xmin <- seq_df$x - 0.42
  seq_df$xmax <- seq_df$x + 0.42
  seq_df
}

wf_df <- make_waterfall_topk_df(
  fcbud_waterfall,
  term1 = "F23", term2 = "F24",
  top_k = 5,               # try 5, 8, 12, etc.
  order_by = "abs_diff",   # or "pct" / "diff"
  verbose = TRUE
)

# Custom order vector
desired_order <- c("F23", "FT21/FT22 (S5)", "TR22/TR23 (S3)", "FT20/FT21 (S7)",
                   "TR20/TR21 (S7)", "TR23/TR24 (S1)", "Other", "F24")

# Ensure labels match those in wf_df$label exactly
wf_df$label <- factor(wf_df$label, levels = desired_order)
wf_df <- wf_df[order(as.numeric(wf_df$label)), ]

plot_ly(
  wf_df,
  type = "waterfall",
  x = ~label,
  y = ~value,
  measure = ~ifelse(type == "Start", "absolute",
                    ifelse(type == "End", "total", "relative")),
  text = ~comma(round(value)),
  textposition = "outside",
  connector = list(line = list(color = "rgba(0,0,0,0.3)", width = 1.5)),
  decreasing = list(marker = list(color = "#d95f02")),
  increasing = list(marker = list(color = "#1b9e77")),
  totals     = list(marker = list(color = "#6b6b6b"))
) %>%
  layout(
    title = "FCB upper-division headcount<br><sup>Cohort contributions to change F23 → F24</sup>",
    yaxis = list(title = "Headcount", range = c(3000, NA)),
    xaxis = list(title = ""),
    showlegend = FALSE
  )

```

# Decomposition visual

```{r}

# Assumes fcbud_waterfall has: Cohort, Semester, Diff, A,B,C,D,E, Pct_of_Total
# Pick the same rows you plotted earlier (e.g., Top-K by |Diff|)
top_k <- 5
wf_long <- fcbud_waterfall %>%
  filter(Cohort != "Total") %>%
  arrange(desc(abs(Diff))) %>%
  slice_head(n = top_k) %>%
  mutate(Label = paste0(Cohort, " (S", Semester, ")")) %>%
  select(Label, Cohort, Semester, Diff, `Admissions`=A, `% FCB`=B, `Retention`=C, `Net transfers`=D, `Promotion`=E) %>%
  pivot_longer(cols = c(`Admissions`,`% FCB`,`Retention`,`Net transfers`,`Promotion`),
               names_to = "Factor", values_to = "Contribution") %>%
  # Nice hover text
  mutate(hover = paste0(
    Label, "<br>",
    Factor, ": ", comma(round(Contribution)), "<br>",
    "Total contribution (Δ): ", comma(round(Diff))
  ))

# order x by total absolute contribution, newest on left if you prefer
wf_long <- wf_long %>%
  group_by(Label) %>% mutate(abs_tot = first(abs(Diff))) %>% ungroup() %>%
  mutate(Label = factor(Label, levels = unique(Label[order(-abs_tot)])))

# --- 2) ggplot (your existing stacked chart), just add aes(text=hover) ---
p_stack <- ggplot(wf_long, aes(x = Label, y = Contribution, fill = Factor, text = hover)) +
  geom_col(position = "stack", width = 0.75) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Decomposition of cohort contributions to change F23 → F24",
    subtitle = "Positive (above zero) and negative (below zero)",
    x = NULL, y = "Headcount contribution", fill = "Factors"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 28, hjust = 1)
  )

# --- 3) Make it interactive ---
ggplotly(p_stack, tooltip = "text")

```

# Backcast table

```{r}
# 1. Remove FT25, FT26, TR25, TR26
backcast_table <- backcast_table %>%
  filter(!Cohort %in% c("FT25", "FT26", "TR25", "TR26"))

# 2. Set NA forecasts to zero (all *_forecast columns)
backcast_table <- backcast_table %>%
  mutate(across(ends_with("_forecast"), ~ifelse(is.na(.x), 0, .x)))

# 3. Calculate delta columns if they are not present
if(!"F24_pre_error" %in% names(backcast_table)) {
  backcast_table <- backcast_table %>%
    mutate(
      F24_pre_error  = F24_pre_forecast - F24_pre_actual,
      F24_ud_error   = F24_ud_forecast  - F24_ud_actual,
      S25_pre_error  = S25_pre_forecast - S25_pre_actual,
      S25_ud_error   = S25_ud_forecast  - S25_ud_actual
    )
}

# 4. FT17/TR19: recalculate Δ columns (all <= 0)
backcast_table <- backcast_table %>%
  mutate(
    F24_pre_error  = ifelse(Cohort %in% c("FT17", "TR19"),
                            F24_pre_forecast - F24_pre_actual, F24_pre_error),
    F24_ud_error   = ifelse(Cohort %in% c("FT17", "TR19"),
                            F24_ud_forecast - F24_ud_actual, F24_ud_error),
    S25_pre_error  = ifelse(Cohort %in% c("FT17", "TR19"),
                            S25_pre_forecast - S25_pre_actual, S25_pre_error),
    S25_ud_error   = ifelse(Cohort %in% c("FT17", "TR19"),
                            S25_ud_forecast - S25_ud_actual, S25_ud_error)
  )

# 5. FT24/TR24: model columns NA, Δ columns 0
backcast_table <- backcast_table %>%
  mutate(
    across(ends_with("_forecast"), ~ifelse(Cohort %in% c("FT24", "TR24"), NA_real_, .x)),
    across(ends_with("_error"),    ~ifelse(Cohort %in% c("FT24", "TR24"), 0, .x))
  )

# 6. Totals row (ignore NAs)
# Remove any existing Total row before adding a new one
backcast_table <- backcast_table %>% 
  filter(Cohort != "Total")

# Add the new totals row as before
totals_row <- backcast_table %>%
  summarise(across(where(is.numeric), ~sum(.x, na.rm = TRUE))) %>%
  mutate(Cohort = "Total") %>%
  select(Cohort, everything())

backcast_table <- bind_rows(backcast_table, totals_row)

# 7. Build the gt table with your formatting and vertical lines after each Δ column
backcast_gt <- gt(backcast_table) %>%
  tab_header(
    title = md("**Back-cast F24 Enrollment by Cohort (using S24 info only)**"),
    subtitle = "Forecast vs. actuals for FCB Pre-Major and Upper-Division"
  ) %>%
  cols_label(
    F24_pre_actual    = "F24 Pre-Major (actual)",
    F24_pre_forecast  = "F24 Pre-Major (model)",
    F24_pre_error     = "Δ (model - actual)",
    F24_ud_actual     = "F24 UD (actual)",
    F24_ud_forecast   = "F24 UD (model)",
    F24_ud_error      = "Δ (model - actual)",
    S25_pre_actual    = "S25 Pre-Major (actual)",
    S25_pre_forecast  = "S25 Pre-Major (model)",
    S25_pre_error     = "Δ (model - actual)",
    S25_ud_actual     = "S25 UD (actual)",
    S25_ud_forecast   = "S25 UD (model)",
    S25_ud_error      = "Δ (model - actual)"    
  ) %>%
  fmt_number(
    columns = where(is.numeric),
    decimals = 0,
    sep_mark = ","
  ) %>%
  sub_missing(
    columns = everything(),
    missing_text = ""
  ) %>%
  cols_align(align = "center", columns = everything()) %>%
  tab_options(table.font.size = px(14)) %>%
  opt_row_striping() %>%
  # Add vertical lines after each delta column
  tab_style(
    style = cell_borders(sides = "right", color = "grey60", weight = px(2)),
    locations = list(
      cells_body(columns = "F24_pre_error"),
      cells_body(columns = "F24_ud_error"),
      cells_body(columns = "S25_pre_error")
    )
  )

# 5. 
backcast_gt

```
